---
title: "Seurat_tutorial"
author: "Beatriz Val"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(Seurat)
library(patchwork)
library(presto)
library(ggplot2)
```

### **Load data**

```{r}
srat_hc <- readRDS("Data/Hong2021_HC1_srat.RDS")
srat_hc@assays$RNA@counts[1:5,1:5]
```

\

```{r}
gene_rename_tab = data.frame("original" = rownames(srat_hc[["RNA"]]),
                               "new" = sapply(strsplit(rownames(srat_hc[["RNA"]]), "-"), 
                                              function(x) paste0(x[-1], collapse = "-")))
  gene_rename_tab$new[gene_rename_tab$new==""] = gene_rename_tab$original[gene_rename_tab$new==""]
  
  counts_gn = DelayedArray::rowsum(srat_hc[["RNA"]]@counts, group = gene_rename_tab$new)
  colnames(counts_gn) = colnames(srat_hc[["RNA"]])
  
  srat_hc = CreateSeuratObject(counts = counts_gn, meta.data = srat_hc@meta.data)
```

## Standard pre-processing workflow

### QC and selecting cells for further analysis

Compute mitochondrial QC metrics

```{r}
srat_hc[["percent.mt"]] <- PercentageFeatureSet(object = srat_hc, pattern = '^MT-')
head(srat_hc)
```

```{r}
# Visualize QC metrics as a violin plot
plot <- VlnPlot(srat_hc, features = c("nFeature_RNA")) + ggtitle('RNA Feature') + xlab(NULL) + theme(legend.position = 'none', axis.text.x = element_blank())
plot1 <- VlnPlot(srat_hc, features = c("nCount_RNA")) + ggtitle('RNA Count') + xlab(NULL) + theme(legend.position = 'none', axis.text.x = element_blank())
plot2 <- VlnPlot(srat_hc, features = c("percent.mt")) + ggtitle('Mitochondrial Percentage') + xlab(NULL) + theme(legend.position = 'none', axis.text.x = element_blank())

plots <- plot+plot1+plot2 + plot_layout(ncol = 3, widths = 4, heights = 7, axes = 'collect')

ggsave(filename = 'Final_figs/Vln_hc.jpg', height = 10, width =16 , plot = plots, quality = 50)
plots
```

Cut-offs: 200, 2500 and 20

```{r}
# FeatureScatter is typically used to visualize feature-feature relationships, but can be used for anything calculated by the object, i.e. columns in object metadata, PC scores etc.

plot1 <- FeatureScatter(srat_hc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(srat_hc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

```{r}
#srat_hc <- subset(srat_hc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
srat_hc <- srat_hc[,srat_hc$nFeature_RNA > 200 & srat_hc$nFeature_RNA < 2500 & srat_hc$percent.mt < 20]
```

```{r}
saveRDS(srat_hc, file = "srat_hc.rds")
```

## Normalizing the data

```{r}
srat_hc <- NormalizeData(srat_hc, normalization.method = "LogNormalize", scale.factor = 10000)
```

### Find Variable Features

```{r}
srat_hc<- FindVariableFeatures(srat_hc, selection.method = "vst", nfeatures = 2000)

srat_hc$orig.ident
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(srat_hc), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(srat_hc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2
```

### Scaling the data

```{r}
all.genes <- rownames(srat_hc)
srat_hc <- ScaleData(srat_hc, features = all.genes)


```

### Linear Dimensional Reduction - PCA

```{r}
srat_hc <- RunPCA(srat_hc, features = VariableFeatures(object = srat_hc))
```

```{r}
# Examine and visualize PCA results a few different ways
print(srat_hc[["pca"]], dims = 1:5, nfeatures = 5)
```

```{r}
VizDimLoadings(srat_hc, dims = 1:2, reduction = "pca")
```

```{r}
DimPlot(srat_hc, reduction = "pca") + NoLegend()
```

```{r}
DimHeatmap(srat_hc, dims = 1, cells = 500, balanced = TRUE)
```

```{r}
DimHeatmap(srat_hc, dims = 1:15, cells = 500, balanced = TRUE)
```

```{r}
elbowplot <- ElbowPlot(srat_hc, ndims = 50) + ggtitle('Dimensionality') + theme(plot.title = element_text(hjust = 0.5, vjust = 1.5))
ggsave(filename = 'Final_figs/elbowplot.jpg', height = 7, width = 12, plot = elbowplot, quality = 50)
elbowplot
```

Cutoff - 15

## Cluster the cells

```{r}
srat_hc <- FindNeighbors(srat_hc, dims = 1:15)
srat_hc <- FindClusters(srat_hc, resolution = 0.5)
head(Idents(srat_hc), 5)
```

## Run non-linear dimensional reduction (UMAP/tSNE)

```{r}
srat_hc <- RunUMAP(srat_hc, dims = 1:15)
# note that you can set `label = TRUE` or use the LabelClusters function to help label individual clusters
dimplot <- DimPlot(srat_hc, reduction = "umap", label = T) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  theme(axis.title = element_text(size = 18), axis.text = element_blank(), axis.ticks = element_blank(), legend.text = element_text(size = 18)) +
  guides(colour = guide_legend(override.aes = list(size = 10)))


ggsave(filename = 'Final_figs/clusters_hc.jpg', height = 7, width = 12, plot = dimplot, quality = 50)
dimplot
```

```{r}
saveRDS(srat_hc, file = "umap.rds")
```

## Annotations

### Finding deferentially expressed features (cluster biomarkers)

```{r}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(srat_hc, ident.1 = 2)
head(cluster2.markers, n = 5)
```

```{r}
# find all markers distinguishing cluster 5 from clusters 0 and 3
cluster5.markers <- FindMarkers(srat_hc, ident.1 = 5, ident.2 = c(0, 3))
head(cluster5.markers, n = 5)
```

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive
# ones
srat_hc.markers <- FindAllMarkers(srat_hc, only.pos = TRUE)
srat_hc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1)

View(srat_hc.markers)
```

```{r}
cluster8.markers <- FindMarkers(srat_hc, ident.1 = 8, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)
cluster8.markers
```

```{r}
VlnPlot(srat_hc, features = c("MS4A1", "CD79A"))
```

```{r}
# you can plot raw counts as well
VlnPlot(srat_hc, features = c("NKG7", "PF4"), slot = "counts", log = TRUE)
```

```{r}
genes = c("MS4A1", "GNLY", "CD3E", "CD4", "CD8A")

featureplot <- list()


for (gene in genes){
plot <- FeaturePlot(srat_hc, features = gene ) +
  scale_color_gradient(low = 'lightgrey', high = '#4B0082', limits = c(0, 6))+
  guides(color = "none") +
  xlab(NULL) +
  ylab(NULL)
if (length(featureplot) != 0 && length(featureplot) != 3) {
  plot <-  plot +
  theme_void() + 
  theme(plot.title = element_text(face = 'bold', hjust = 0.5, vjust = 0.5))
  }

featureplot[[gene]] <- plot
}



featureplot[[1]] <- featureplot[[1]] +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  xlab('UMAP 1') +
  ylab('UMAP 2')

featureplot[[4]] <- featureplot[[4]] + 
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  xlab('UMAP 1') +
  ylab('UMAP 2')

featureplot[['LYZ']] <- FeaturePlot(srat_hc, features = 'LYZ' ) +
  xlab(NULL) +
  ylab(NULL) +
  theme_void() +
  theme(plot.title = element_text(face = 'bold', hjust = 0.5, vjust = 0.5)) +
  scale_color_gradient(low = 'lightgrey', high = '#4B0082', limits = c(0, 6))
  



featureplots <- featureplot[[1]] +
  featureplot[[2]] +
  featureplot[[3]] +
  featureplot[[4]] +
  featureplot[[5]] +
  featureplot[[6]] +
  plot_layout(ncol = 3, nrow = 2, widths = 6, heights = 4, axes = 'collect_x') 


ggsave(filename = 'Final_figs/markers_for_cell_type_hc.jpg', height = 7, width = 12, plot = featureplots, quality = 50)
featureplots
```

```{r}
srat_hc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(srat_hc, features = top10$gene) + NoLegend()
```

```{r}
new.cluster.ids <- c("CD4", "CD4", "CD8", "Mono-Mac", "NK", "B cells", "CD8", "Platelets", "Endothelial Cells", 'Mono-Mac', 'Mast cells_Basophils')
names(new.cluster.ids) <- levels(srat_hc)
srat_hc <- RenameIdents(srat_hc, new.cluster.ids)
DimPlot(srat_hc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```

```{r}
plot <- DimPlot(srat_hc, reduction = "umap", label = TRUE, repel = T, label.size = 4.5) +
  xlab("UMAP 1") +
  ylab("UMAP 2") +
  theme(axis.title = element_text(size = 18), axis.text = element_blank(), axis.ticks = element_blank(), legend.text = element_text(size = 18)) +
  guides(colour = guide_legend(override.aes = list(size = 10)))


ggsave(filename = "Final_figs/Annotations_hc.jpg", height = 7, width = 12, plot = plot, quality = 50)
```

```{r}
srat_hc[['Ident']] <- Idents(srat_hc)
```

```{r}
saveRDS(srat_hc, file = "srat_hc_final.rds")
```

Create info to put in chatGPT to help define the clusters

```{r}
annotationPrompt = function(presto_df, n = 10, 
                            pval_thr = 0.05, pct_diff = 20, logFC_thr = 0.2,
                            add_info = NULL, print.it = T){
  require(dplyr)
  
  topmk = presto_df |>
    group_by(cluster) |> # for each cluster
    filter(p_val_adj<=pval_thr) |> # only p-value below 0.05
    filter(pct.1>(pct.2+pct_diff)) |> 
    filter(avg_log2FC>logFC_thr) |> 
    arrange(-avg_log2FC) |>
    top_n(n = n, wt = avg_log2FC) # top genes per cluster, ranked by logFC
  
  prompt = "I have performed a scRNA-seq analysis, and have encountered various clusters, for which I calculated the marker genes. The top marker genes for each cluster are as follows:\n"
  for(cl in unique(topmk$cluster)){
    prompt = paste0(prompt, " - cluster ", cl, ": ", 
                    paste0(topmk$gene[topmk$cluster==cl], collapse = ", "), "\n")
  }
  
  if(!is.null(add_info)){
    prompt = paste0(prompt, 
                    "\nTo interpret these clusters, these additional sample characteristics should be considered: ", 
                    paste0(sapply(names(add_info), function(x) paste0(x, ": ", add_info[x])), 
                           collapse = "; "), "\n")
  }
  
  prompt = paste0(prompt, 
                  "Can you tell me what are the most likely cell types that each cluster matches to? And report the top 5 genes used to reach this annotation.")
  
  if(print.it){
    cat(prompt)
  }
  return(prompt)
}
```

```{r}
annotationPrompt(srat_hc.markers, n = 25, pct_diff = 0.2, add_info = list('tissue' = 'blood'))
```
